---
title: "Danish monarchs"
author: "Henriette Stenius Wohlin"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document:
    toc: TRUE
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Danish Monarchs and set up

This is an R markdown document for my assignment about danish monarchs for week 47. 


```{r Setup}
library(tidyverse)
library(here)
```

```{r load-data}
monarchs <- read.csv2("data/danish_monarchs.csv")
```


## Data fixing

I had some problems with sorting my monarchs so I had to Google how to do it.
Rstudio would sort my x-axis alphabetically and not in order of my data.

To change this I had to:

```{r Fixing data, echo=TRUE}
#Turn your 'treatment' column into a character vector
monarchs$monarch. <- as.character(monarchs$monarch.)
#Then turn it back into a factor with the levels in the correct order
monarchs$monarch. <- factor(monarchs$monarch., levels=unique(monarchs$monarch.))
```

Source: https://stackoverflow.com/questions/12774210/how-do-you-specifically-order-ggplot2-x-axis-instead-of-alphabetical-order 

This code have to before the following code in order for it to work

## Visualize the data

```{r Danish monarchs and reign over time, echo=TRUE}
monarchs %>% 
  select(monarch., reign_length) %>% 
  ggplot(aes(x = monarch., y = reign_length)) +
  geom_col() +
  labs(title = "Danish monarchs reign in years",
       y = "Years",
       x = "Monarch") +
  scale_y_continuous(breaks=seq(0,60,5)) +
  theme(axis.text.x = element_text(angle = 90))
```
