---
title: "W48"
author: "Henriette Stenius Wohlin"
date: "2022-12-04, updated `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

```{r libraries, warning=FALSE, message=FALSE}
library(rvest)
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
library(tidyverse)
library(forcats)
library(scales)
```

## Scrape the data

First we scrape data from website: 
```{r url}
url <- "https://www.worldometers.info/world-population/population-by-country/"
# scrape the website
url_html <- read_html(url)
```
-Code from Adela Sobotkova, https://github.com/Digital-Methods-HASS/WebscrapingPoliceKillings/blob/master/scraping.Rmd

Now we can extract the whole HTML table:

```{r scrape-table}
whole_table <- url_html %>% 
 html_nodes("table") %>%
 html_table()  #str(whole_table) turns out to be a list
str(whole_table)
whole_table[[1]]
whole_table
```
-Code from Adela Sobotkova, https://github.com/Digital-Methods-HASS/WebscrapingPoliceKillings/blob/master/scraping.Rmd

Now we make the data more readable in a new_table, or so we can visualize it better:

```{r html-to-df}
new_table <- do.call(cbind,unlist(whole_table, recursive = FALSE)) 
head(new_table)
```
-Code from Adela Sobotkova, https://github.com/Digital-Methods-HASS/WebscrapingPoliceKillings/blob/master/scraping.Rmd

I want to rename columm 2 and 3, so it has no ( ) in them:
```{r}
colnames(new_table)[2] = "Country"
colnames(new_table)[3] = "Population"
```
Source: https://sparkbyexamples.com/r-programming/rename-column-in-r/

The dataset is currently a character matrix. I will change it to tibble so we can clean it up
```{r clean-data}
new_table <- as_tibble(new_table)
str(new_table)
```
However the Population columm is still a character vector and has , between the numbers. This means we can't change the vector. So we use this code:
```{r}
col_conv <-  c("Population")                                             

my_data_updated <-  new_table                                            
my_data_updated[ , col_conv] <-  lapply(my_data_updated[ , col_conv], 
                                       function(x){ as.numeric(as.character(gsub(",", "", x))) })
```
Source: https://www.youtube.com/watch?v=FjhEVnWmvOs

Now we can change Population from a character vector to a numeric one:
```{r}
data <- my_data_updated %>% 
	mutate(Population = as.numeric(Population))
```
-Code from Adela Sobotkova, https://github.com/Digital-Methods-HASS/WebscrapingPoliceKillings/blob/master/scraping.Rmd

## Combaring population between countries

Now I want to try and compare the population of Denmark, Norway, Finland and Sweden:
```{r plot-world, eval=TRUE}
my_data_updated %>% 
  mutate(Country = fct_reorder(Country, desc(Population))) %>%
    group_by(Country, 
           Population) %>% 
  filter(Country %in% c("Norway", "Sweden", "Finland", "Denmark")) %>%
ggplot(aes(x = Country, y = Population, color = Country)) +
geom_col() +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6))
```

-Code from Adela Sobotkova, https://github.com/Digital-Methods-HASS/WebscrapingPoliceKillings/blob/master/scraping.Rmd

Source for scale command: https://www.roelpeters.be/scale-ggplot-y-axis-millions-or-thousands-r/





